---
title: "Nucleic Acids"
---

<!-- Google tag (gtag.js)-->


```{=html}
<script>
window.dataLayer = window.dataLayer || [];
function gtag(){dataLayer.push(arguments);}
gtag('js', new Date());

gtag('config', 'G-S1L73VGCG5');
</script>
```

| The difference between the composition of the **DNA** (deoxyribonucleic acid)\index{DNA} molecule and that of **RNA** (ribonucleic acid)\index{RNA} lies in two main aspects. The first concerns the coverage of its five constituent *monomers*, called *heterocyclic bases*, whether [adenine](https://chemapps.stolaf.edu/jmol/jmol.php?model=adenine&script=Console){target="_blank"}\index{adenine} or [guanine](https://chemapps.stolaf.edu/jmol/jmol.php?model=guanine&script=Console){target="_blank"}\index{guanine}, together forming the *purines*, or [cytosine](https://chemapps.stolaf.edu/jmol/jmol.php?model=cytosine&script=Console){target="_blank"}\index{cytosine}, [thymine](https://chemapps.stolaf.edu/jmol/jmol.php?model=thymine&script=Console){target="_blank"}\index{thymine} and [uracil](https://chemapps.stolaf.edu/jmol/jmol.php?model=uracil&script=Console){target="_blank"}\index{uracil}, together forming *pyrimidines*. In RNA, *uracil* replaces *thymine* in DNA.

| The combination of *bases* with a molecule of ribose and inorganic phosphate gives rise to **nucleosides**. \index{nucleosides}. Exemplifying for [adenosine](https://chemapps.stolaf.edu/jmol/jmol.php?model=adenosine&script=Console){target="_blank"}, a nitrogenous heterocyclic base\index{adenosine} in an N-glycosidic bond\index{N-glycosidic bond} with the hydroxyl linked to carbon-1 of a ribose (RNA, presence of -OH group at C-2) or deoxyribose (DNA) esterified to a phosphate group. In the same way, [guanosine](https://chemapps.stolaf.edu/jmol/jmol.php?model=guanosine&script=Console){target="_blank"}\index{guanosine} originates, [cytidine](https://chemapps.stolaf.edu/jmol/jmol.php?model=cytidine&script=Console){target="_blank"}\index{cytidine}, [thymidine](https://chemapps.stolaf.edu/jmol/jmol.php?model=thymidinescript=Console){target="_blank"}\index{thymidine}, and [uridine](https://chemapps.stolaf.edu/jmol/jmol.php?model=uridine&script=Console){target="_blank"}\index{uridine}.

| By assembling interconnected nucleosides, the structure of the *biopolymer (DNA or RNA)* will be formed by the phosphoester bond between the hydroxyl linked to C5 of the pentose and C3 of the next pentose. The polymeric structure of double-stranded nucleic acids is also maintained by hydrogen bond pairing between *G* and *C* (three bonds) and between *A* and *T* (two bonds), as follows.
\

## B-DNA

| To study the biopolymer, download and load the model of a dodecanucleotide of [B-DNA](./figures/423D.png){target="_blank"}\index{B-DNA} in *JSmol/Jmol*. Remove the water and show the nucleotide sequence of the double-stranded DNA:

```{r, eval=FALSE}
delete water
show sequence
```

| Now represent it as space filled only, and identify the strands separately:

```{r, eval=FALSE}
spacefill only
select :A; color red;select: B; color blue
```

| Next, visualize the interior of the double strand and observe the stacking of the DNA base pairs:

```{r, eval=FALSE}
select all;wireframe only;wireframe 50
select sidechain;color magenta;select backbone;color blue
```

| Also note that the base pairs are in a horizontal plane, and also perpendicular to the axis of the helix. Note the proximity of the *van der Waals* interactions:

```{r, eval=FALSE}
dots on
```

| Restrict the view to just one base pair, zoom in on it, and notice that the bases induce the slight twisting of the helix, which is important to the structure of DNA, although not identifiable in the **Watson-Crick model**\index{Watson-Crick model}.

```{r, eval=FALSE}
restrict 8 or 7
zoom 0
```

| Also notice the ribose residues, and see that the plane of the residues is perpendicular to the axis of the helix, with no coplanarity between them.

```{r, eval=FALSE}
select backbone;color cpk
```

| Which atom of the *furanose ring* rests above and below its plane? To do this, rotate the molecule so that the phosphate groups of the backbone are on the left, and the plane of the furanose ring is slightly tilted. Note that there is a phosphate bonding below at *C3'*, as well as above at *C5'*, forming a **3'-5' phosphodiester bond**\index{3'-5' phosphodiester bond}.

| To visualize the *hydrogen bonds* of a single base pair, rotate the model after the commands that follow.

```{r, eval=FALSE}
restrict 8 or 17
zoom 0
calculate hbonds
```

| And to visualize the hydrogen bonds of the entire structure:

```{r, eval=FALSE}
select all;
wireframe only;
wireframe 40;
zoom 0;
calculate hbonds
```

| Note that this visualization allows us to verify the *stability of the DNA strand*\index{DNA stability} corroborated by its H-bonds in addition to *base stacking* (hydrophobic)\index{base stacking.

### DNA Clefts

| Inspect the **DNA clefts**\index{DNA clefts}. Note that the *major* and *minor* clefts are defined on the paper plane in a didactic way, from the overlap of the two strands (twist point). To do this, position the double helix perpendicularly and rotate it, to notice that this twist point "walks" through the structure:

```{r, eval=FALSE}
select all;
trace only;
spin 10
```

| By definition, the major cleft is always next to the base pairs, while the minor cleft is always next to the furanoses.

## A-DNA

| Download and load the [A-DNA](./figures/adna.png){target="_blank"}\index{A-DNA} model into *Jmol/JSmol*. Visualize the single strand of this DNA:

```{r, eval=FALSE}
delete water;
select all;
wireframe only;
wireframe 70
```

| Identify the bases and riboses:

```{r, eval=FALSE}
select sidechain;
color magenta;
select backbone;
color lightblue
```

| Now observe the A-DNA molecule seen from above, rotating the image or typing:

```{r, eval=FALSE}
rotate x 90 # 90 degree rotation on the X axis
```

| Note that there is a channel inside the molecule, which is not present in the B-DNA structure.

## Z-DNA

| Download and load the [Z-DNA](./figures/zdna.png){target="_blank"}\index{Z-DNA} structure into *JSmol/Jmol*. This is a biological construct containing Z-DNA, a structure characteristically rich in GC pairs. To highlight Z-DNA only:

```{r, eval=FALSE}
restrict dna;
show sequence;
wireframe only;
wireframe 40
```

| Rotate the molecule and observe that the double strand is formed by a hexanucleotide (GC)$_{3}$.

| Also observe the intercalation of cytidine and guanosine nucleosides in the structure:

```{r, eval=FALSE}
select g:A; color red;
select c:A; color blue
```

| The term Z-DNA comes from the [zigzag aspect](./figures/zigzag.png){target="_blank"}\index{Z-DNA} of the orientation of the riboses. Note that from a ribose in blue (below), the phosphodiester bond appears to the left of the phosphorus atom, going up to the other ribose molecule in red, the set forming the zigzag aspect.

## Interaction of nucleic acids with ligands

| The formation of nucleic acid-ligand complexes will be illustrated by the interaction of a *double-stranded RNA* with *ethidium*, using the code DRB018. As you can see, this is an identification external to the PDB. To download this structure, you can visit the [Nucleic Acids Database](http://ndbserver.rutgers.edu){target="_blank"}\index{NAD - database of acids nucleic acid}. But you can also load the structure from the [ligand-nucleic acid complex](./figures/dnacomplex.png){target="_blank"}\index{ligand-nucleic acid complex}.

| If you are on the [Nucleic Acids Database](http://ndbserver.rutgers.edu){target="_blank"} website, enter the code *DRB018* in the search window. Download the resulting coordinate file in the lower corner of the window, under "Assymmetric Unit coordinates". The file, in the *gz* attribute, can be opened in *Jmol.*

| Then load the file and render it appropriately:

```{r, eval=FALSE}
delete water;
select all;
wireframe only;
wireframe 70
```

| Now rotate the molecule and notice its intercalation between the base stacking of the double-stranded RNA.

## Nucleosome

| Load the model [1aoi](./figures/1aoi.png){target="_blank"}\index{nucleosome} into *Jmol/JSmol*. It is the central particle of a **nucleosome** composed of an octamer of *histones*\index{histones} wrapped in two superhelical turns of double-stranded DNA.

| Better identify the structures contained in the nucleosome, separating the protein part from the nucleic part, and rotate the complex to observe its characteristics (DNA-protein interaction envelope).

```{r, eval=FALSE}
select protein;cartoon only;color structure
select nucleic;wireframe only;color cpk
```

| Histones interact with DNA mainly through electrostatic interactions with its phosphate groups. To visualize this interaction:

```{r, eval=FALSE}
select protein;backbone only;color lightgrey;
select positive; wireframe 80;color green
```

## Transfer RNA

| Like proteins, **transfer RNA**\index{t-RNA} also has primary, secondary and tertiary structures, first evidenced by X-ray diffractometry in 1972. To learn a little more, download and load the [t-RNA](./figures/trna.png){target="_blank"}\index{t-RNA} model, and rotate the structure to see its three-dimensional "L" shape (or *cloverleaf*, as mentioned in the books). Zoom in and render the structure, outline the coloring from the blue 5' to the red 3' terminal, rotate the structure and notice the helical character of the tRNA:

```{r, eval=FALSE}
delete water;trace only;color group
zoom 0; zoom 200;wireframe only;wireframe 70
```

| Position the ends on the right (5' and 3' ends), and observe the *loops* that form the **secondary structure**\index{t-RNA secondary structure}, the *anticodon loop*\index{anticodon} (below), the *D loop* on the left, and the *T and C loops* on the right.

| More specifically, type the commands below to better visualize the *loops*:

```{r, eval=FALSE}
select 44-48;color red # variable loop
select 10-13 or 22-25;color yellow # stem D
```

| Now observe the parallelism and stacking of the bases along the t-RNA stems:

```{r, eval=FALSE}
wireframe only;
wireframe 80;
select backbone;
wireframe
```

| Also select the three anticodon residues for *Phe* of t-RNA by:

```{r, eval=FALSE}
select 36-38; color red
```

| The structure of t-RNA is maintained stable primarily by *Watson-Crick* base pairing, as well as by H-bonds. To observe this:

```{r, eval=FALSE}
select all;
color structure;
calculate hbonds;
hbonds 0.2;
color hbonds black
```

| Stability is also enhanced by H-bond interactions between ribose residues. To observe this:

```{r, eval=FALSE}
select nucleic;
wireframe 80;
color green;
restrict backbone;
select(6-8,48-50);
color cpk
```

| Bring the sugar residues closer and notice (by mouse click) that *2'-OH* of ribose 7 can form an *H* bond with *O* no. 49 of the sugar ring. The length of an H-bond lies between 2 and 4 Angstrom.

| To check the potential for forming an *H* bond between the carbohydrates above, right-click, select "Measurements" (or its translation, if you are in your native language) and "Click for distance measurement". Now click on the first desired atom (2'-OH) and then on the one that has the potential to form the pairing (O-49). And remember that 1 nm is equal to 10 Angstrom.

| In general, the base-base, base-sugar, base-phosphate, sugar-sugar, and sugar-phosphate pairings contribute as *H* bonds to the stability of t-RNA.

| Now visualize *a Watson-Crick pairing*, albeit with the *G* and *C* residues in non-helical portions of the t-RNA, by typing:

```{r, eval=FALSE}
select all;color lightgrey;wireframe;hbonds off # model cleaning
select 19 or 56;color red
```

| To observe the importance of base interactions for the **tertiary structure**\index{t-RNA tertiary structure} of the t-RNA type:

```{r, eval=FALSE}
hbonds off;backbone only;
color structure;
select 18 or 55;color orange;
wireframe 100;
hbonds on;hbonds 0.3
```

| tRNA interacts with amino acid transporters by **triplets**\index{triplets} of codons in its structure (or trinucleotides). Highlighting the GGC triplet:

```{r, eval=FALSE}
select all;hbonds off;wireframe only;
color lightgrey;select (9,13,22,46);wireframe 80;
color cpk
```

## Ribosomes

| For a structural study of **ribosomes**, download and load the [30S subunit](./figures/30S.png){target="_blank"}\index{30S subunit} of the ribosome of *Thermus thermophilus*. Distinguish the nucleic from the protein structures:

```{r, eval=FALSE}
cartoon only;select nucleic;color green; select protein;color orange
```

| Is the structure represented "more RNA" or "more protein"? Try to visualize it by rotating the model by:

```{r, eval=FALSE}
select all;spacefill only
```

| Is the protein structure unique? To solve this question, rotate the structure from:

```{r, eval=FALSE}
select all;select nucleic;color cpk;
select protein;color chains
spin 10
```

| And how does the nucleic part fold?

```{r, eval=FALSE}
restric rna
```

| Observe the spherical shape and the nuclear characteristic of the RNA, to which the various proteins that form the 30S subunit are complexed.

### 70S ribosomal particle

| To observe it, download and load into *Jmol/JSmol* the model of the [70S particle](./figures/70S.png){target="_blank"}\index{70S subunit} of *Thermus thermophilus* complexed to m-RNA and t-RNA. Alternatively, you can download the original file *4v42* from *Jmol* itself. But note that if you try to do it through *Menu*, *Get PDB*, you will not be successful. For this model, you need to download the file from the PDB website, in *PDBx/mmCIF Format (gz)* format. This is necessary because it is a file subdivided into two structures, 30S and 50S subunits. In addition to these, the bacterial ribosome has a hexanucleotide segment of m-RNA complexed with three t-RNAs.

| The model shows three t-RNAs inside (blue-red), a large set of proteins (dark gray), and the remaining nucleic structure (orange). The 70S ribosome is quite complex, with several sites, both for interaction with t-RNA (A, P and E) and for interaction with protein subunits. To find out how many of these, type *show info*, and you will see that the structure has 49 polypeptide chains.  At the terminal part of t-RNAs, the nearby A and P sites define the site of peptidyl transferase activity on the ribosome.